---
title: "Getting started with dqcheckrGUI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dqcheckrGUI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

`dqcheckrGUI` is a point-and-click Shiny interface for
[dqcheckr](https://github.com/mickmioduszewski/dqcheckr). It lets you
configure dataset quality checks, run them against incoming file deliveries,
and browse historical results — without writing any R code.

## Prerequisites

Install `dqcheckr` before installing `dqcheckrGUI`:

```{r}
install.packages("dqcheckr")
devtools::install_github("mickmioduszewski/dqcheckrGUI")
```

## Launching the app

```{r}
dqcheckrGUI::run_app()
```

The app opens in your default browser. It reads and writes configuration files
from the working directory you launched it from, so set that to your project
folder first.

To use the example Star Wars configuration files included with the package, copy
them to your working directory:

```{r}
file.copy(
  system.file("extdata/example_config", package = "dqcheckrGUI"),
  "config",
  recursive = TRUE
)
```

## The interface

The sidebar has four sections:

- **Datasets** — list of configured datasets; click one to view or edit it
- **▶ Run** — run a quality check against a new file delivery
- **⏱ History** — browse past runs and compare any two deliveries
- **⚙ Global Config** — set default thresholds and file paths

---

## Setting up a dataset (8-step wizard)

Click **+ New dataset** in the sidebar to open the configuration wizard.

### Step 1 — Dataset identity

Give the dataset a short machine-readable name (letters, numbers, and
underscores only, starting with a letter) and an optional description.

### Step 2 — File location

Choose how files are identified:

- **Folder scan** — point to a folder; the app picks up the most recently
  modified file on each run. Suitable for delivery processes that drop files
  into a fixed directory.
- **Explicit file paths** — list one or more specific file paths. Useful when
  files are versioned by name (e.g. `dataset_v1.csv`, `dataset_v2.csv`).

### Step 3 — Format and structure

The app shows a raw preview of the file. Select the format:

- **CSV** — set the delimiter, quote character, and whether the file has a
  header row.
- **Fixed-width (FWF)** — drag the column boundary ruler over the preview to
  define column positions. The ruler snaps to character positions.

A parsed preview updates as you adjust settings.

### Step 4 — Column classification

Review the detected columns. For each column you can:

- **Set a type override** — force a column to be treated as `character` even
  if it looks numeric (e.g. postcodes, phone numbers, BSB codes).
- **Mark as key column** — used in uniqueness checks.
- **Mark as expected** — columns not present in a delivery will raise a
  schema warning.

### Step 5 — Column rules

Expand any column to add optional per-column validation rules such as a
maximum missing rate or a required value pattern.

### Step 6 — Rule overrides

Override the global default thresholds for this dataset only. For example,
raise the maximum missing rate for a column that is known to be sparsely
populated, or tighten the row-count change threshold for a stable reference
dataset. Leave fields unchanged to inherit from Global Config.

### Step 7 — Custom checks (optional)

Point to a plain R file that defines a `custom_checks(df)` function. The
function receives the dataset as a data frame and returns a list of
`dq_result()` objects. See the `dqcheckr` documentation for the function
signature.

Leave this field blank to skip custom checks.

### Step 8 — Review and save

A summary of all settings is shown alongside a YAML preview of the
configuration that will be written. Click **Save config ✓** to write the file.

---

## Running a quality check

Click **▶ Run** in the sidebar.

1. Select a dataset from the dropdown.
2. Click **▶ Run check**.

The app runs `dqcheckr::run_dq_check()` in a background process so the
interface remains responsive. Progress is shown in the log area. When the
check completes, the HTML report opens automatically in a new browser tab.

---

## Browsing history

Click **⏱ History** in the sidebar to see a table of all past runs across
all datasets, showing the date, dataset name, status (PASS / WARN / FAIL),
row count, and report link.

To compare two deliveries:

1. Tick the checkbox next to exactly two runs **from the same dataset**.
2. Click **Compare drift ▶**.

The drift report opens in a new tab, showing column-by-column changes between
the two snapshots.

---

## Global configuration

Click **⚙ Global Config** to set defaults that apply to all datasets unless
overridden at the dataset level.

**Infrastructure paths**

- **Snapshot database** — path to the SQLite file where run summaries and
  column statistics are stored (default: `data/snapshots.sqlite`).
- **Report output directory** — where HTML reports are written (default:
  `reports/`).

**Default rule thresholds**

| Setting | Default | Meaning |
|---|---|---|
| Max missing rate | 0.05 | Flag a column if more than 5 % of values are blank |
| Max non-numeric rate | 0.01 | Flag a numeric column if more than 1 % of values cannot be parsed |
| Min row count | 0 (disabled) | Fail if the delivery has fewer rows than this |
| Max row count change | 10 % | Warn if row count changes by more than 10 % vs previous delivery |
| Max mean shift | 20 % | Warn if a numeric column mean shifts by more than 20 % |
| Max missing change | 2 pp | Warn if missing rate changes by more than 2 percentage points |
| Max non-numeric change | 1 pp | Warn if non-numeric rate changes by more than 1 percentage point |
| Type inference threshold | 0.90 | A column is typed numeric if ≥ 90 % of values parse as numbers |

**Schema change flags**

Enable or disable warnings for: new columns appearing, columns being dropped,
column types changing, or column order changing between deliveries.

Click **Save global config** to write the changes.

---

## Learn more

See `vignette("dqcheckr", package = "dqcheckr")` for a full description of
every quality check (QC, SC, and CP series), per-column configuration options,
custom checks, and the snapshot database schema.
